---
title: "Number of functions"
author: "Amelia McNamara"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
```

This file shows how I generated the datasets `allfunctions.csv` and `csfunctions.csv`, which enumerate all the functions I showed in my labs in the Fall 2020 semester. 

Because parts of this script need to be run in different folders, this document cannot be run as-is. 

## Step 1: Get all the R code

The first step in this process is to process all RMarkdown documents into .R script files. The `purl()` function will do this for you. I moved this file into the folder where all my lab files were kept.

```{r, eval = FALSE}
# this stuff gets run in the labs folder
here::i_am("LL labs/number_of_functions.Rmd")
library(here)
lab_filenames <- tibble(filename = list.files(path = here("LL labs"), recursive = TRUE, full.names = TRUE))
lab_filenames <- lab_filenames %>%
  filter(str_detect(filename, ".Rmd")) %>%
  filter(str_detect(filename, "PreLab"))

map(lab_filenames$filename, purl)
```

After running this code, you should have a .R file that corresponds to every RMarkdown document you initially had. (RMarkdown documents are preserved.) Rather than moving the files programmatically or attempting to write the appropriate filepath, I drag-and-dropped all these files into my paper folder. 

## Step 2: Parse the code 
Now that you have a bunch of .R files, you can use the function `getParseData()` to parse the files, and find the functions that are used. 

```{r}
# this is the stuff you can run in the papers folder
here::i_am("data/processing_data/number_of_functions.Rmd")
library(here)
purled_labs <- list.files(path = here("data/prelabs"), full.names = TRUE)
names(purled_labs) <- list.files(path = here("data/prelabs"))

parsed_all <- map_dfr(purled_labs, ~getParseData(parse(file = .)), .id = "filename")

allfunctions <- parsed_all %>%
  filter(token == "SYMBOL_FUNCTION_CALL")  %>%
  mutate(section = case_when(str_detect(filename, "tidy") ~ "tidy",
                             str_detect(filename, "formula") ~ "formula",
                             TRUE ~ NA_character_)) 

allfunctions <- allfunctions %>%
  group_by(section, text) %>%
  summarize(n = n())

write_csv(allfunctions, "data/processed/allfunctions.csv")
```

At the end of this process, you should have a .csv file with all the functions from the .R files. 


## Step 3 (optional): Parse other files 

For my classes, I had attempted to generate a cheatsheet a priori that contained all the functions used in the class. So, I wanted to parse these documents separately. This section of the document goes through that process, which is very similar to steps 1 and 2 above. 

```{r}
cs_filenames <- tibble(filename = list.files(path = here("data/cheatsheets"), recursive = TRUE, full.names = TRUE))

map(cs_filenames$filename, purl)
```


```{r}
purled_cs <- list.files(path = here("data/cheatsheets"), full.names = TRUE) 
names(purled_cs) <- list.files(path = here("data/cheatsheets"))
purled_cs <- purled_cs[str_ends(purled_cs, ".R")]

parsed_cs <- map_dfr(purled_cs, ~getParseData(parse(file = .)), .id = "filename")

csfunctions <- parsed_cs %>%
  filter(token == "SYMBOL_FUNCTION_CALL")  %>%
  mutate(section = case_when(str_detect(filename, "tidy") ~ "tidy",
                             str_detect(filename, "formula") ~ "formula",
                             TRUE ~ NA_character_)) 

csfunctions <- csfunctions %>%
  group_by(section, text) %>%
  summarize(n = n())

write_csv(csfunctions, "data/processed/csfunctions.csv")
```



